### 2023-04-11 latex/HP1_TranslComm.tex
new base text

\input{Preamble-Print}
\makepagestyle{HPed}
\makeoddhead{HPed}{\small\texteng{HP Transl. \& Comm.}}{}{\small\texteng{\today}}
\makeevenhead{HPed}{\small\texteng{HP Transl. \& Comm.}}{}{\small\texteng{\today}}
\makeoddfoot{HPed}{}{\small\texteng{\thepage}}{}
\makeevenfoot{HPed}{}{\small\texteng{\thepage}}{}

% additions/changes 2022-03-12 mm; 2022-6-1 MD:
\SetTEIxmlExport{autopar=false}
\NewDocumentEnvironment{translation}{O{}}{\textcolor{blue}{\textbf{Transl.:}}}{}
\NewDocumentEnvironment{philcomm}{O{}}{
	\textcolor{blue}{\textbf{Comm.:}}}{}
\NewDocumentEnvironment{sources}{O{}}{
	\textcolor{blue}{\textbf{Testimonia:}}}{}

\setlength{\parskip}{0.3em}
\setlength\parindent{0pt}
\def\startverse{\begin{ekdverse}\begin{itshape}}
\def\endverse{\end{itshape}\end{ekdverse}}
\setvnum{}

\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

\chapter*{Translation \& philological commentary}
\subsection*{1.1}
\begin{translation}[hp01_001]
Homage to the glorious Ādinātha by whom the knowledge of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
\end{translation}

\begin{sources}[hp01_001]
Yogasārasaṃgraha (line 3090), Gheraṇḍasaṃhitā
\end{sources}

%\begin{philcomm}[hp01_001]
%\end{philcomm}

\subsection*{1.2}
\begin{translation}[hp01_002]
Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the doctrine of Haṭhayoga solely for [attaining] Rājayoga.
\end{translation}

%\begin{philcomm}[hp01_002]
%\end{philcomm}

\subsection*{1.3}
\begin{translation}[hp01_003]
For those who are ignorant of Rājayoga because of confusion in the darkness of many opinions the compassionate Svātmārāma composes the Lamp on Haṭha.
\end{translation}

%\begin{philcomm}[hp01_003]          
%\end{philcomm}

\subsection*{1.4}
\begin{translation}[hp01_004]
In fact, Matsyendra, Gorakṣa and other [siddhas] knew the doctrine of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
\end{translation}

%\begin{philcomm}[hp01_004]          
%\end{philcomm}

@@

%\begin{philcomm}[hp01_005]          
%\end{philcomm}

\subsection*{1.6}
\begin{translation}[hp01_006]
Manthānabhairava, Siddhabuddha, and Kanthaḍi, Pauraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
\end{translation}

%\begin{philcomm}[hp01_006]          
%\end{philcomm}

@@

%\begin{philcomm}[hp01_007]          
%\end{philcomm}

@@

%\begin{philcomm}[hp01_008]          
%\end{philcomm}

\subsection*{1.9}
\begin{translation}[hp01_009]
These great siddhas and others have destroyed the rod of death through the power of Haṭhayoga and wander in the universe.
\end{translation}

%\begin{philcomm}[hp01_009]          
%\end{philcomm}

\subsection*{1.10}
\begin{translation}[hp01_010]
Haṭha is considered a refuge for those who are burnt by the pains of transmigration. Haṭha is the foundational tortoise for the worlds of all yogas.
\end{translation}

%\begin{philcomm}[hp01_010]          
%\end{philcomm}

\subsection*{1.11}
\begin{translation}[hp01_011]
The doctrine of Haṭha should be kept very secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
\end{translation}

\begin{sources}[hp01_011]
ŚS 5.254; BKhP 10v4, YCM
\end{sources}

%\begin{philcomm}[hp01_011]          
%\end{philcomm}

\subsection*{1.12}
\begin{translation}[hp01_012]
In well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live remotely in a small hut.
\end{translation}

\begin{sources}[hp01_012]
GŚ 32cd; BKhP 107v1, YCM
\end{sources}

%\begin{philcomm}[hp01_012]          
%\end{philcomm}

\subsection*{1.13}
\begin{translation}[hp01_013]
It has a small door, without cracks and holes, its length is not too high or low, thickly smeared with cow dung in the proper way, clean, free from everything that annoys, adorned with a pavilion, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
\end{translation}

\begin{sources}[hp01_013]
BKhP 107v3, YCM
 \end{sources}

\begin{philcomm}[hp01_013]          
\emph{piṭharaṃ} good
\end{philcomm}

\subsection*{1.14}
\begin{translation}[hp01_014]
Locating oneself in a hut of such a kind, free from all worry, one should constantly practise yoga in the way taught by one's guru.
\end{translation}

@@

\subsection*{1.15}
\begin{translation}[hp01_015]
Overeating, exertion, chatter (gossiping/bickering?), sticking to rules, associating with people, inconstancy: through [these] six, yoga will be abandoned.
\end{translation}

\begin{philcomm}[hp01_015]          
Impossible to decide on meaning of \emph{niyamāgraha}, avagraha invisible, Jyotsnā takes it as over-insistence (as if \emph{āgraha} was implied) as he relates it to extreme ascetic practice; will be abandoned if \emph{prahāsyate} is adopted.
\end{philcomm}

\subsection*{1.16}
\begin{translation}[hp01_016]
From zeal, conviction, resolve, understanding of the truth [of yoga], discernment, abandonment of associating with people: by [these] six, on the other hand, yoga is successful.
\end{translation}

\begin{philcomm}[hp01_016]          
\emph{tattva} can sometimes refer to the practices of yoga: e.g., \emph{tritattva} in AS 13.12, 14.2--3.
\end{philcomm}

\subsection*{1.17}
\begin{translation}[hp01_017]
Because it is the first auxiliary of haṭha, āsana is taught first. This (\emph{tad}) āsana brings about steadiness, good health and dexterity.
\end{translation}

%\begin{philcomm}[hp01_017]          
%\end{philcomm}

\subsection*{1.18}
\begin{translation}[hp01_018]
I shall now teach some of the postures which have been accepted by sages (\emph{munis}) such as Vasiṣṭha and yogis such as Matsyendra.
\end{translation}

%\begin{philcomm}[hp01_018]          
%\end{philcomm}

\subsection*{1.19}
\begin{translation}[hp01_019]
Correctly placing the soles of both feet between the knees and thighs [and] sitting up with the body straight: they call that the auspicious [pose].
\end{translation}

\begin{sources}[hp01_019]
VS 1.68 etc
\end{sources}

%\begin{philcomm}[hp01_019]          
%\end{philcomm}

\subsection*{1.20}
\begin{translation}[hp01_020]
One should place one's right heel on the left, at the side of the [lower] back, and on the right the left in the same way. This is the Cow's Mouth [posture]; it is like a cow's mouth.
\end{translation}

\begin{sources}[hp01_020]
VS 1.70, YY? 3.5; BKhP 93v7
\end{sources}

%\begin{philcomm}[hp01_020]          
%\end{philcomm}

\subsection*{1.21}
\begin{translation}[hp01_021]
Having placed one foot on one thigh and the seated thigh on the other foot. Vīrāsana is taught [thus]. 
\end{translation}

\subsection*{1.22}
\begin{translation}[hp01_022]
Having blocked the anus with the crossed ankles, focused [the yogi] remains [thus]. Adepts of yoga know that this is kūrmāsana.

Or: While focused, [the yogi] should bind the anus with crossed ankles. This is kūrmāsana. Experts in yoga know this.
\end{translation}

\subsection*{1.23}
\begin{translation}[hp01_023]
Having assumed padmāsana, inserted the hands between the knees and thighs, and placed [the hands] on the ground, [the yogi] remains in the air. This is kurkuṭāsana

Or: [The yogi] should assume padmāsana, insert his hands between the knees and thighs onto the ground and place himself in the air. This is kurkuṭāsana.
\end{translation}

\begin{philcomm}[hp01_023]
Kurkuṭa/kurkkuṭa variant in \getsiglum{V1,J10ac,V3} is attested in Pañcatantra (M-W).
\end{philcomm}

\subsection*{1.24}
\begin{translation}[hp01_024]
Bound in kurkuṭāsana, [the yogi] binds the neck with the hands and lies supine like a tortoise. This is uttānakūrmāsana. 
\end{translation}

\begin{philcomm}[hp01_024]
\emph{°bandhasthaḥ} or \emph{°vat kṛtvā}? Only V1 has latter, which is simpler. Are the others trying to improve it? Stemmatically ambiguous as \emph{°bandha°} is on one branch (V3/J8, V19) and \emph{°madhya°} the other (J10, J17, N17). V1 looks like an outlier.

Adopt \emph{bandhastho} with note. \emph{°bandhastha} not found in any other texts. \emph{°vat kṛtvā} is not good.
\end{philcomm}

\subsection*{1.25}
\begin{translation}[hp01_025]
Having clasped the big toes with hands, [the yogi] should perform a bow drawing action as far as the ear. This is called dhanurāsana.
\end{translation}

\begin{philcomm}[hp01_025]
At a stretch (!), this could mean drawing a crossbow, with both hands taken back to the ears, but probably just means like a normal bow.

>>>>

\input{Preamble-Print}
\input{CommStyle}
\begin{document}
\pagestyle{HPed}
\begin{ekdosis}
\SetLineation{lineation = none,}

\chapter*{Translation \& philological commentary}
\subsection*{1.1}
\begin{translation}[hp01_001]
Homage to the glorious Ādinātha by whom the system of Haṭhayoga was taught. It shines forth like a ladder for one desirous of climbing to the lofty terrace of the royal palace.
\end{translation}

\begin{testimonia}[hp01_001]
Cf. Yogasārasaṃgraha, p. 54.
\startverse
sadādināthāya namo'stu tubhyaṃ yenopadiṣṭā haṭhayogavidyā | \\
virājate pronnatarājayogam āroḍhum icchoradhirohiṇīva ||
\endverse

Gheraṇḍasaṃhitā 1.1
\startverse
ādīśvarāya praṇamāmi tasmai yenopadiṣṭā haṭhayogavidyā | \\
virājate pronnatarājayogam āroḍhum icchor adhirohiṇīva ||
\endverse
\end{testimonia}

\begin{philcomm}[hp01_001]
The reading \emph{rājasaudha} is preferable for poetical reasons, as explained in Hanneder 2020, p. 128–130. Also, the \emph{Jyotsnā} explains that the \emph{alaṃkāra} here is a comparison (\emph{upamā}), which consists of four elements: (1) a particle expressing the comparison
(\emph{iva}), (2) the object compared (\emph{upameya}), i.e., the \emph{haṭhayogavidyā} leading to Rājayoga (\emph{rājayogaprāpikā}), (3) the property compared, that is, effortlessness (\emph{anāyāsena}) and, most importantly, (4) an image to compare with, which in this case is \emph{rājasaudha}. This idea is confirmed by Brahmānanda, when he sums up and completes the poetical details as: \emph{yathā pronnatasaudham āroḍhum icchor adhirohiṇy anāyāsena saudhaprāpikā bhavati evaṃ haṭhadīpikāpi pronnatarājayogam āroḍhum icchor anāyāsena rājayogaprāpikā bhavatīti upamālaṅkāraḥ}.

The Sanskrit poeticians explain that when some elements of a comparison remains unexpressed we get an incomplete (\emph{lupta}) comparison. Often words like \emph{iva} or \emph{yathā} are missing or the common property, but if we read \emph{rājayoga} we lose the \emph{upamāna} (i.e., \emph{rājasaudha}). One part of the comparison should not be missing. But from the perspective of \emph{alaṃkāraśāstra} the verse has a problem that has surely caused the dilemma: it either leaves the \emph{upameya} or the \emph{upamāna} incomplete. Brahmānanda has mentioned both \emph{yathā adhirohiṇī saudhaprāpikā bhavati evaṃ haṭhadīpikā rājayogaprāpikā bhavati}, but had to supply \emph{rājasaudha}, while his text reads \emph{rājayoga}. But to have the \emph{upameya} in the text is odd. Even Brahmānanda could only know of the image (\emph{rājasaudha}) from the alternative reading he did not accept. However, it seems likely that the author would have included the \emph{upamāna} in the text, as in our critical text, so that the reader would know that the upper terrace of the palace is an image for \emph{rājayoga}. It is further likely that the substitution of \emph{rājayoga} for \emph{rājasaudha} was the result of the tendency to insert the word \emph{yoga} in the opening verses of the text wherever possible, even where it does not fit, as can be seen in 1.2d (\emph{haṭhayogopadiśyate}) and 1.3b (\emph{rājayogam ajānatām}). In the latter case the poetical image has also been lost.

The metre of 1.1 is Indravajrā
\end{philcomm}

\subsection*{1.2}
\begin{translation}[hp01_002]
Having bowed to the glorious guru, the Lord, the yogi Svātmārāma has taught the system of Haṭhayoga solely for [attaining] Rājayoga.
\end{translation}

\begin{philcomm}[hp01_002]
The reading \emph{yogopadiśyate} (J7,J10,N17,W4, etc.) is only possible if one accepts that double \emph{sandhi} is a feature of the style of composition, which it is not. Moreover, this reading appears to have resulted from an attempt to replace the word \emph{vidyā} with \emph{yoga} in the opening verses of the text.
\end{philcomm}

\subsection*{1.3}
\begin{translation}[hp01_003]
The compassionate Svātmārāma holds the Lamp on Haṭha for those who are ignorant of the royal path because of wandering in the darkness of many opinions.
\end{translation}

\begin{testimonia}[hp01_003]
Haṭharatnāvalī 1.4
\startverse
bhrāntyā bahumatadhvānte rājayogam ajānatām |\\
kevalaṃ rājayogāya haṭhavidyopadiśyate || 
\endverse
\end{testimonia}

\begin{philcomm}[hp01_003] 
Most witnesses have \emph{rājayogam ajānatām} (`for those ignorant of Rājayoga') in 1.3b. While this reading is well attested by the manuscripts transmission of the \emph{Haṭhapradīpikā} and some manuscripts of the \emph{Haṭharatnāvalī} (note that P,T,t1 of the critical edition [Gharote 2009: 5 n. 2] have \emph{rājamārgam ajānataḥ}), the reading of a royal path (\emph{rājamārgam}) is consistent with the metaphor of people wandering from a path in the darkness and, therefore, most probably authorial.

In 1.1d both \emph{kṛpākaraḥ} and \emph{kṣamākaraḥ} are attested. The former is preserved by the V1 and group 2 whereas the latter by some manuscripts of the delta group. Since the context is the author helping yogis who have strayed from the royal path, \emph{kṛpākara} makes better sense. As Brahmānanda notes, this compound can be understood as one who is compassionate (\emph{kṛpā} + \emph{kara}) or one who is a mine (i.e., a rich source) of compassion (\emph{kṛpā} + \emph{ākara}). In the Devanagari transmission, the \emph{kṣa} of \emph{kṣamākaraḥ} may have arisen as a mistake for \emph{kṛ}.       
\end{philcomm}

\subsection*{1.4}
\begin{translation}[hp01_004]
In fact, Matsyendra, Gorakṣa and other [perfected yogis] knew the system of Haṭha, and the yogi Svātmārāma knows it owing to their favour.
\end{translation}

\begin{testimonia}[hp01_004]
Post-15th Century\\
Haṭharatnāvali 1.3
\startverse
haṭhavidyāṃ hi gorakṣamatsyendrādyā vijānate |\\
ātmārāmo 'pi jānīte śrīnivāsas tathā svayam || 
\endverse
\end{testimonia}

\begin{philcomm}[hp01_004]   
The word \emph{athavā} (‘or’) is difficult to construe here. Brahmānanda understands it as conjunction (\emph{athavāśabdaḥ samuccaye}), and this is how we have interpreted it. The variant \emph{mahāyogī} in group 4c (G5, J4, J11, Ko, P15) avoids this difficulty, but is probably an attempt to remove the difficulty of understanding \emph{athavā}. One could emend to \emph{tathā} but this would be a bold intervention given the weight of evidence supporting \emph{’thavā}.   
\end{philcomm}

@@

\begin{testimonia}[hp01_005]
Haṭharatnāvalī 1.80
\startverse
śrīādināthamatsyendraśābarānandabhairavāḥ |\\
śāraṅgīmīnagorakṣavirūpākṣabileśayāḥ || 
\endverse
\end{testimonia}

\begin{philcomm}[hp01_005]     
In Śaiva texts which predate the haṭha corpus, Mīnanātha and Matsyendra are one and the same, but they are differentiated in later Tibetan and Indian lists of siddhas (Mallinson 2019:273 n.35).   
\end{philcomm}

\subsection*{1.6}
\begin{translation}[hp01_006]
Manthānabhairava, Siddhabuddha, and Kanthaḍi, Koraṇṭaka, Surānanda, Siddhapāda, Carpaṭi.
\end{translation}

\begin{testimonia}[hp01_006]
Haṭharatnāvalī 1.81
\startverse
manthānabhairavo yogī siddhabuddhaś ca kandalī |\\
korandakaḥ surānandaḥ siddhipādaś ca carpaṭī ||
\endverse

Caturbhuja Misra's Mugdhāvabodhinī (1.7.8) on the Rasahṛdayatantra
\startverse
manthānabhairavo yogī siddhabuddhaś ca kanthaḍī |\\
koraṇṭakaḥ surānandaḥ siddhapādaś ca carpaṭī ||
\endverse
\end{testimonia}

\begin{philcomm}[hp01_006]        
Many manuscripts preserve the reading \emph{pauraṇṭaka}. We have accepted \emph{koraṇṭaka} because this name is associated with the \emph{Haṭhābhyāsapaddhati}. This reading is well attested by various manuscripts of the \emph{Haṭhapradīpikā}, including those of group 2, as well as those of the \emph{Haṭharatnāvalī} (Gharote 2009: 35 n. 3), which also attest \emph{gonandaka}. Also, \emph{Goraṇṭakuḍu} is the name of a disciple of Gorakṣanātha in the \emph{Navanāthacaritramu} (Jones 2017:194 n.3). The compound \emph{siddhapāda} could be a respectful affix. However, it seems unlikely here because it would cross the \emph{pāda} break.  
\end{philcomm}

@@

\begin{testimonia}[hp01_007]
Haṭharatnāvalī 1.82
\startverse
karoṭiḥ pūjyapādaś ca nityanātho nirañjanaḥ |\\
kapālī bindunāthaś ca kākacaṇḍīśvarāhvayaḥ || 
\endverse

Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
​​\startverse
kaṇerī pūjyapādaśca nityanātho nirañjanaḥ |\\
kapālī bindunāthaśca kākacaṇḍīśvaro gajaḥ |
\endverse
\end{testimonia}

\begin{philcomm}[hp01_007]   
It is possible that \emph{pūjyapāda} could be a respectful affix to the name Kāṇerī. The variant \emph{dhvaninātha} may have resulted from a transposition of the first two syllables of \emph{nityanātha}.    
\end{philcomm}

@@

\begin{testimonia}[hp01_008]
Haṭharatnāvalī 1.83
\startverse
allamaḥ prabhudevaś ca naiṭacūṭiś ca ṭiṇṭiṇiḥ |\\
bhālukir nāgabodhaś ca khaṇḍakāpālikas tathā || 
\endverse

Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
\startverse
āllamaḥ prabhudevaś ca ghoḍācolī ca ṭhiṇṭhinī |\\
bhālukir nāgadevaś ca khaṇḍī kāpālikas tathā ||
\endverse
\end{testimonia}

\begin{philcomm}[hp01_008]  
The name Allamaprabhudeva (sometimes Allama Prabhu Deva or Allama Prabhudeva in secondary literaure) is largely transmitted as allamaḥ prabhudevaḥ, as though it were two names, although some manuscripts of the \emph{Haṭhapradīpikā} have \emph{allamaprabhudevaś ca} (i.e., Tue,V3,V8,V13,V16,V22,Vu,N24,N26) and so do some of the \emph{Haṭharatnāvalī} (i.e., P,T,t1 in Gharote 2009: 35 n. 8).

The names Nāgabodha, Nāgabodhi, Naradeva, Nāgadeva all seem possible in 1.8c.

Many witnesses have khaṇḍa and kāpālika as separate names. However, khaṇḍakāpālika is well attested. Examples include \emph{Kathāsaritsāgara} 121.5 ff. (check), \emph{Bṛhatkathāmañjarī} 10.45 (check) and Vajrapāṇi’s \emph{Laghutantraṭīkā}, p.45 (\emph{vīrāḥ khaṇḍakāpālikādayaś caturviṃśatiḥ}). It may be a derogatory name for a Kāpālika, coined perhaps by an outsider and connoting something like a defective Kāpālika in the sense of a `part-time' Kāpālika. Alternatively, it could simply refer to one who used a broken skull as a bowl.
\end{philcomm}

\subsection*{1.9}
\begin{translation}[hp01_009]
Having destroyed the rod of death through the power of Haṭhayoga, these great perfected yogis and others wander in the world.
\end{translation}

\begin{testimonia}[hp01_009]
Haṭharatnāvalī 1.84
\startverse
ityādayo mahāsiddhāḥ haṭhayogaprasādataḥ |\\
khaṇḍayitvā kāladaṇḍaṃ brahmāṇḍe vicaranti te ||
\endverse

Caturbhuja Misra's Mugdhāvabodhinī on the Rasahṛdayatantra
\startverse
ityādayo mahāsiddhā rasabhogaprasādataḥ |\\
khaṇḍayitvā kāladaṇḍaṃ trilokyāṃ vicaranti te |
\endverse

Haṭhatattvakaumudī 17.24
\startverse
ūrdhvaṃretaḥprabhāvena sanakādyā maharṣayaḥ |\\
khaṇḍayitvā kāladaṇḍaṃ yathecchaṃ viharanti te || 24 ||
\endverse
\end{testimonia}

\begin{philcomm}[hp01_009]        
The reference to \emph{brahmāṇḍa} (‘the world’) implies liberation-in-life (\emph{jīvanmukti}) and physical immortality.  
\end{philcomm}

\subsection*{1.10}
\begin{translation}[hp01_010]
Haṭha is considered a refuge for those who are burnt by the scorching torment of transmigration. Haṭha is the tortoise that supports the worlds of all yogas.
\end{translation}

\begin{testimonia}[hp01_010]
Yogasārasaṃgraha (YSS), p.53.
\startverse
saṃsāratāpataptānāṃ samāśrayahaṭho haṭhaḥ |\\
aśeṣayogajagatām ādhārakamaṭho haṭhaḥ ||
\endverse
\end{testimonia}

\begin{philcomm}[hp01_010] 
The compound \emph{saṃsāratāpa°} is well attested and found elsewhere (e.g., \emph{Viṣṇupurāṇa} 6.7.62, \emph{Agnipurāṇa} 371.1, \emph{Haṭhatattvakaumudī} 38.92, emph{Haṭhābhyāsapaddhati} ms. 46/440, f. 1v). The reading of \emph{samāśrayo} in V1 is metrically faulty. The accepted reading \emph{āśrayo [']yam} is better metically and well attested. N23 preserves the reading \emph{samāśrayamaṭho haṭhaḥ} (similar to the \emph{Yogasārasaṃgraha}), which is similar to the fourth \emph{pāda}. Though this reading is possible, it seems more likely that it came about through dittography. The word \emph{°jagatāṃ} is probably authorial because it makes good sense with  \emph{ādhārakamaṭha} in light of the cosmological notion that the tortoise supports all the worlds. However, this reading may not have been understood by some and was changed in the vulgate and other witnesses to \emph{°yuktānāṃ} instead.      
\end{philcomm}

\subsection*{1.11}
\begin{translation}[hp01_011]
The doctrine of Haṭha should be kept completely secret by those yogis who are desiring success. When it is secret it becomes potent. However, when it has been revealed, it becomes impotent.
\end{translation}

\begin{sources}[hp01_011]
Śivasaṃhitā
\startverse
haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\% (ab missing in some ŚS witnesses)
bhaved vīryavatī guptā nirvīryā ca prakāśitā || 5.254
\endverse
\end{sources}

\begin{testimonia}[hp01_011]
Yogacintāmaṇi
\startverse
tathā haṭhapradīpikāyām—\\
haṭhavidyā paraṃ gopyā yoginā siddhim icchatā |\\
bhaved vīryavatī guptā nirvīryā tu prakāśiteti ||
\endverse

BKhP 10v4
\end{testimonia}

\begin{philcomm}[hp01_011]        
Either the singular or plural of yogin could be read here. The singular is well attested among the testimonia. However, the weight of the manuscript evidence favours the plural.  
\end{philcomm}

\subsection*{1.12}
\begin{translation}[hp01_012]
In a well-ruled, righteous region, with plenty of food and free of disturbances, the Haṭhayogi should live remotely in a small hut.
\end{translation}

\begin{sources}[hp01_012]
References to maṭhikā here.
\end{sources}

\begin{testimonia}[hp01_012]
Haṭharatnāvalī 1.66
\startverse
surāṣṭre dhārmike deśe subhikṣe nirupadrave |\\
ekāntamaṭhikāmadhye sthātavyaṃ haṭhayoginā ||
\endverse

Yogacintāmaṇi
\startverse
haṭhapradīpikāyām—\\
surājye dhārmike deśe subhikṣe nirupadrave |\\
ekānte maṭhikāmadhye sthātavyaṃ haṭhayoginā ||
\endverse

BKhP 107v1
\end{testimonia}

\begin{philcomm}[hp01_012]        
In the \emph{Jyotsnā} and printed editions of the \emph{Haṭhapradīpikā}, including one by Digambara and Kokaje (1970: 6), this verse has the additional hemistich, \emph{dhanuḥpramāṇaparyantaṃ śilāgnijalavarjite}. This hemistich derives from the \emph{Gorakṣaśataka} (32cd), which has \emph{°paryante} instead of \emph{°paryantaṃ}. It stipulates that the hut should be built in a place measuring up to a bow length and free from rocks, fire and water. None of the early manuscripts have this hemistich which suggests that it was added at a later time. Nonetheless, it appears in over a dozen manuscripts that were consulted for this edition. These manuscripts are not close to an early hyparchetype of the text.
\end{philcomm}

\subsection*{1.13}
\begin{translation}[hp01_013]
It has a small door and is without cracks, holes and potsherds. It extends not too high or low, and is thickly smeared with cow dung in the proper way. It is clean, free from everything that annoys, furnished on the outside with a verandah, altar and well, surrounded by a wall: these are the characteristics of the yoga hut as taught by the adept practitioners of haṭha.
\end{translation}

\begin{sources}[hp01_013]
Cf. Dattātreyayogaśāstra
\startverse
suśobhanaṃ maṭhaṃ kuryāt sūkṣmadvāraṃ tu nirvraṇaṃ || 54 ||\\
suṣṭhu liptaṃ gomayena sudhayā vā prayatnataḥ |\\
matkuṇair maśakair bhūtair varjitaṃ ca prayatnataḥ || 55 ||\\
dine dine susammṛṣṭaṃ sammārjanyā hy atandritaḥ |\\
vāsitaṃ ca sugandhena dhūpitaṃ guggulādibhiḥ || 56 ||\\
malamūtrādibhir vargair aṣṭādaśabhir eva ca |\\
varjitaṃ dvārasampannaṃ vastrāvaraṇam eva vā || 57 ||
\endverse

\end{sources}

\begin{testimonia}[hp01_013]
Suśruta 6.17.67:
\emph{gṛhe nirābādhe}

Yogacintāmaṇi
\startverse
alpadvāram arandhragartaghaṭitaṃ nāpy uccanīcāyitam |\\
samyaggomayasāndraliptavimalaṃ niḥśeṣajantūjjhitam |\\
bāhye maṇḍapakūpavediracitaṃ prākārasaṃveṣṭitam |\\
proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||
\endverse

Haṭharatnāvalī 1.67
\startverse
alpadvāram arandhragartapiṭharaṃ nātyuccanīcāyataṃ\\
samyaggomayasāndraliptavimalaṃ niḥśeṣabādhojjhitaṃ |\\
bāhye maṇḍapavedikūparuciraṃ prākārasaṃveṣṭitam\\
proktaṃ yogamaṭhasya lakṣaṇam idaṃ siddhair haṭhābhyāsibhiḥ ||
\endverse

BKhP 107v3
\end{testimonia}

\begin{philcomm}[hp01_013]  
The syntax of this verse is rather problematic. One would expect the features of the hut, which are listed in the first three pādas of the verse, to be in the nominative case, with the words \emph{idaṃ lakṣaṇaṃ} in the fourth pāda referring back to them. However, the compounds in the first three pādas appear to qualify \emph{lakṣaṇa} as though they were adjectives, and this appears to have been the way the verse was composed.

The manuscripts preserve many different readings at the end of the compound beginning with \emph{arandhragarta°}. The reading \emph{°piṭharaṃ} has been suggested by Dominic Goodall, who understands it as potsherds (ask ref from Dominic?). The sense is that the hut should be free of rubbish, such as potsherds. One would expect a word for a defect in a hut that is similar to, but not the same as, cracks (\emph{randhra}) and holes (\emph{garta}). For this reason, the reading \emph{°vivaraṃ} looks like a patch, as its meaning does not add anything to \emph{°randhragarta°}. The reading \emph{viṭapaṃ} is well attested among the witnesses of the \emph{Haṭhapradīpikā} and \emph{Haṭharatnāvalī}. However, it is difficult to construe in this context unless it was intended to refer to creepers or branches that might invade or encroach upon the hut.

Nearly all the early mss. have \emph{°bādhojjitaṃ}, whereas the Yogacintāmaṇi and V19 have the more easily understood reading of \emph{°jantūjjhitaṃ} (‘free from creatures’). However, \emph{°bādhojjitaṃ} may have been original because it is so well attested and a hut is described similarly in \emph{Suśrutasaṃhitā} 6.17.67 (\emph{gṛhe nirābādhe}).
\end{philcomm}

\subsection*{1.14}
\begin{translation}[hp01_014]
Locating oneself in a hut of such a kind, free from all worry, [the yogi] should practise only yoga, in the way taught by his guru.
\end{translation}

\begin{sources}[hp01_014]
Cf. Amanaska 2.15
\startverse
evaṃvidhaṃ guruṃ labdhvā sarvacintāvivarjitaḥ\\
sthitvā manohare deśe yogam eva samabhyaset
\endverse
\end{sources}

\begin{testimonia}[hp01_014]
Yogacintāmaṇi
\startverse
evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\
gurūpadiṣṭamārgeṇa yogam eva sadābhyased iti ||
\endverse

Haṭharatnāvalī 1.68
\startverse
evaṃvidhe maṭhe sthitvā sarvacintāvivarjitaḥ |\\
gurūpadiṣṭamārgeṇa yogam eva sadābhyaset ||
\endverse
\end{testimonia}

@@

\subsection*{1.15}
\begin{translation}[hp01_015]
Overeating, exertion, idle chatter, not sticking to rules, socialising, sensuality: through [these] six, yoga is lost.
\end{translation}

\begin{testimonia}[hp01_015]
Yogacintāmaṇi
\startverse
atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\
janasaṅgaś ca laulyaṃ ca ṣaḍbhir yogaḥ praṇaśyati ||
\endverse

Haṭharatnāvalī 1.77
\startverse
atyāhāraḥ prayāsaś ca prajalpo niyamagrahaḥ |\\
janasaṅgaṃ ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||
\endverse

Yuktabhavadeva 4.25 (attributed to the śivayoga)
\startverse
atyāhāraḥ prayāsaśca prajalpo niyamāgrahaḥ |\\
janasaṃgaś ca laulyaṃ ca ṣaḍbhir yogo vinaśyati ||
\endverse

Jyotsnā
\startverse
śītodakena prātaḥsnānanaktabhojanaphalāhārādirūpaniyamasya grahaṇaṃ niyamagrahaḥ |
\endverse

Yogaprakāśikā
\startverse
niyamāgrahaḥ vakṣyamāṇaniyamāparipālanaṃ
\endverse
\end{testimonia}

\begin{philcomm}[hp01_015]        
It is impossible to be certain about the meaning of \emph{niyamagraha}, as the manuscripts do not indicate whether an \emph{avagraha} (i.e., \emph{prajalpo ’niyamagraha}) was intended. Although \emph{yama} and \emph{niyama} are not included in the \emph{Haṭhapradīpikā} as auxiliaries of Haṭhayoga, verse 2.14 implies that \emph{niyama} is necessary at least in the early stages of establishing a practice. Furthermore, verse 3.82 suggests that a yogi who does not practice \emph{niyama} might obtain success in yoga through the practice of \textit{vajroli}. Ambiguity over the role of \emph{yama} and \emph{niyama} in Haṭhayoga appears to have prompted some to insert verses on ten \emph{yama}s and ten \emph{niyama}s after the next verse (1.16). The additional verses may derive from the \emph{Śāradātilakatantra} (25.7–8) or the  \emph{Vasiṣṭhasaṃhitā} (1.38, 1.53). In the \emph{Jyotsnā}, Brahmānanda reads \emph{niyamāgraha} and takes it as though \emph{āgraha} was implied, which yields the meaning of ‘over-insistence on rules’, and he relates it to extreme ascetic practice.
\end{philcomm}

\subsection*{1.16}
\begin{translation}[hp01_016]
However, from zeal, conviction, resolve, contentment, understanding of the truth, and avoiding contact with people, yoga is successful because of [these] six.
\end{translation}

\begin{sources}[hp01_016]
Dharmaputrikā 137cd–38ab%Narharināth ed.?
\startverse
utsāho niścayo dhairyaṃ santoṣas tattvadarśanam | \\
kratūnāṃ copasaṃhāraḥ ṣaṭsādhanam iti smṛtam |
\endverse

Śivadharmottara 10 (W 122r):%JM: what ed or ms is this?
\startverse
utsāhān niścayād dhairyāt santoṣāt tattvadarśanāt |\\
muner janapadatyāgād ṣaḍbhir yogaḥ prasiddhyati |
\endverse

Jñānārṇava 20.1
\startverse
utsāhān niścayād dhairyāt saṃtoṣāt tattvaniścayāt |\\
muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 1 ||
\endverse

Yogabindu by Haribhadra
\startverse
utsāhān niścayād dhairyāt saṃtopāt tattvadarśanāt |\\
muner janapadatyāgāt ṣaḍbhir yogaḥ prasidhyati || 411 ||
\endverse
\end{sources}

\begin{testimonia}[hp01_016]
Yogacintāmaṇi
\startverse
utsāhāt sāhasād dhairyāt tatvajñānād viniścayāt |\\
janasaṅgaparityāgāt ṣaḍbhir yogaḥ prasidhyati ||
\endverse

Haṭharatnāvalī 1.78:
\startverse
utsāhān niścayād dhairyāt tattvajñānārthadarśanāt
[utsāhān niścalād- P,T]\\
bindusthairyān mitāhārāj janasaṅgavivarjanāt |\\
nidrātyāgāj jitaśvāsāt pīṭhasthairyād anālasāt\\
gurvācāryaprasādāc ca ebhir yogas tu sidhyati || 
\endverse
\end{testimonia}
